---
title: "README"
author: "Kim Dancy"
date: "4/13/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r}
library(ggplot2)
closures_recent<-read.csv("~/Desktop/Stat 479/Final-Project/Data/closures_recent")
 closures_recent<-mutate(closures_recent, level2=factor(level, levels=c("Certificate", "Associate's", "Bachelor's", "Other")), ordered=TRUE)
  

p<-ggplot(closures_recent, mapping=aes(x=discharged_borrowers, y=average_discharge)) 
p + geom_point(aes(color=type, size=Enrollment), alpha=0.5) + facet_grid(~level2, margins=TRUE) + scale_x_log10() + scale_y_continuous(labels=scales::dollar)  + theme(legend.position = "bottom", legend.box="vertical") +
 labs(title="Figure 1: Student Borrowers Receiving Closed School Discharge", x="Number of Borrowers Receiving Loan Forgiveness (Log Scale)", y="Average Amount of Debt Forgiven") +
  scale_color_discrete(name="Institutional Control",labels=c("Private Nonprofit", "Private For-Profit", "Public"))


```

_Figure 1_ uses data from the U.S. Department of Education's Federal Student Aid Data Center to show the number of borrowers who received a student loan discharge in the event of an unexpected closure at their higher educaiton institution, along with the amount of debt forgiveness received per borrower. 

The school closure data is merged with infromation from the College Scorecard from the last year in which a school was operating. 

From this merged data, Figure 1 also displays selected characteristics about the institutions where those borrowers were enrolled. 

The figure shows that borrowers who received forgiveness were overwhelmingly enrolled in for-profit institutions, while a number of private non-profit institutions at the four-year level have also closed. 

Borrowers are also concentrated at institutions who primarily award Certificates, and these students typically receive smaller amounts of forigveness compared to those at institutions who primarily award Associate's or Bachelor's degrees. 

Because the number of borrowers in most cases is quite small but contians a few very high values, a log transformation on the x axis helps to better show the variation in the underlying data. 

